Associations between Chlorophyll *a* and various Microcystin-LR Health Advisory

### 2015-12-02 manuscript/manuscript.Rmd
working on #4

In this paper we ...
Thus, the goal of this paper is to utilize the National Lakes Assessment (NLA) data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [@usepa2009national].  So that others may repeat or adjust this analysis, the data, code, and this manuscript are freely available via [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

@@

##Data
We used the 2007 NLA water quality and microcystin-LR concentration data [@usepa2009national]. These data represent a snapshot of water quality from the summer of 2007 and data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.

@@

A detailed discussion of CPA is beyond the scope of this paper; see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  The 95% confidence intervals were calculated from 1000 bootstrapped samples. To identify chlorophyll *a* concentrations of concern we used a 50% conditional probability of exceeding each health advisory level and extracted the minimum chlorophyll *a* concentration that was associated with an upper confidence level of 50% or greater.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.

>>>>

In this paper we build on past efforts and utilize the National Lakes Assessment (NLA) data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [@usepa2009national; @usepa2015drinking; @chorus1999toxic].  We add to past studies by exploring associations with newly announced advisory levels and by also applying a different method, conditional probability analysis.  Utilizing different methods strengthens the evidence for suggested chlorophyll *a* levels that are associated with increased risk of exceeding the health advisory levels as those levels are not predicated on a single analytical method.  So that others may repeat or adjust this analysis, the data, code, and this manuscript are freely available via [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

@@

##Data
We used the 2007 NLA water quality and microcystin-LR concentration data [@usepa2009national]. These data represent a snapshot of water quality from the summer of 2007 for the conterminous United States. Data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.

@@

Conditional probability analysis provides information about the probability of observing one event given another event has also occured.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  We expect to find higher chlorohpyll *a* concentrations to be associated with higher probabilities of exceeding the microcystin-LR health advisory levels.  We also caclulated bootstrapped 95% confidence intervals (CI) using 1000 bootstrapped samples.  Thus, to identify chlorophyll *a* concentrations of concern we identfiy the value of the upper 95% CI across a range of conditional probabilities of exceeding each health advisory level.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.  A more detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.
